Fitting the Undetected: Light-curve Feature Estimation in the Presence of Non-detections

### 2013-03-27 tex/ms.tex
notes to self
typesetting

% Note the status (in preparation, MNRAS submitted date, ApJ accepted date, in press etc).
\slugcomment{DRAFT 2013-03-26 / in preparation}

\begin{abstract}
In multi-epoch imaging surveys, faint variable sources will not be
detectable at all epochs.  In principle, the best approach is to
perform photometric measurements (forced photometry) anyway.
%
In practice, most surveys only provide detection upper limits at some
epochs; in some not even upper limits but just the ``null''
information that the star was not detected.
%
Here we demonstrate with real data on periodic variables from the
All-Sky Automated Survey (ASAS) that these null observations can
in fact be very useful in fitting and parameter estimation.
%
We build a novel light-curve model that includes the per-epoch
detection thresholds as a set of latent parameters, one per data
point; the best-fit model is the maximum-likelihood model after
marginalizing out the latent thresholds.
%
The key assumption is that the detection thresholds are drawn from a
stationary distribution (the parameters of which are free to vary).
%
We demonstrate that this censored-data model outperforms traditional
fitting---in which null observations are discarded from the
fits---in estimating both periods and amplitudes for stars subject to
heavy censoring.
%
In particular, our model eradicates a failure mode in which
traditional fitting obtains an integer fraction of the correct period.
\end{abstract}

@@

In this contribution we look at improving catalog-level time-domain astronomy in the presence of \emph{censored data}.
The problem we consider is that of period and amplitude modeling for large-amplitude variable stars in a data set
in which \emph{no information is reported at all} at the epochs at which the star is not significantly detected.
Furthermore, we will imagine that we also have \emph{no information at all} about the sensitivity of the telescope
at the time of each non-detection; that is, we have no upper limits and we can't reconstruct them from any extant meta-data.
The non-detections are literally database ``null'' values.
This set-up may seem artificial, but it is in fact the situation we find ourselves in for various surveys,
including the ASAS survey, data from which we will model below.
What we will show is that \emph{the ``null'' non-detections are in fact very informative};
amplitude and period fitting will be wrong in general if the non-detections are ignored.

@@

\hogg{either merge in or delete the material below this line}

>>>>

% Note the status (in preparation, MNRAS submitted date, ApJ accepted date, in press etc).
\slugcomment{DRAFT 2013-03-27 / in preparation}

\begin{abstract}
In multi-epoch imaging surveys, faint variable sources will not be
detectable at all epochs.  In principle, the best approach is to
perform photometric measurements (forced photometry) anyway.
%
In practice, most surveys only provide detection upper limits at some
epochs; in some not even upper limits but just the ``null''
information that the star was not detected.
%
Here we demonstrate with real data on periodic variables from the
\textsl{All-Sky Automated Survey} that these null observations can
in fact be very useful in fitting and parameter estimation.
%
We build a novel light-curve model that includes the per-epoch
detection thresholds as a set of latent parameters, one per data
point; the best-fit model is the maximum-likelihood model after
marginalizing out the latent thresholds.
%
The key assumption is that the detection thresholds are drawn from a
stationary distribution (the parameters of which are free to vary).
%
We demonstrate that this censored-data model outperforms traditional
fitting---in which null observations are discarded from the
fits---in estimating both periods and amplitudes for stars subject to
heavy censoring.
%
In particular, our model eradicates a failure mode in which
traditional fitting obtains an integer fraction of the correct period.
\end{abstract}

@@

In this contribution we look at improving catalog-level time-domain astronomy in the presence of \emph{censored data}.
The problem we consider is that of period and amplitude modeling for large-amplitude variable stars in a data set
in which \emph{no information is reported at all} at the epochs at which the star is not significantly detected.
Furthermore, we will imagine that we also have \emph{no information at all} about the sensitivity of the telescope
at the time of each non-detection; that is, we have no upper limits and we can't reconstruct them from any extant meta-data.
The non-detections are literally database ``null'' values.
This set-up may seem artificial, but it is in fact the situation we find ourselves in for various surveys,
including the \textsl{ASAS} survey (CITATION), data from which we will model below.
What we will show is that \emph{the ``null'' non-detections are in fact very informative};
amplitude and period fitting will be wrong in general if the non-detections are ignored.

@@

\hogg{note to self: either merge in or delete the material below this line}
